# Code to Extract Temprature data from Berkeley data set
# Ran it all in chunks since it allows to run multipple lines of code in a row by just clicking the green run button at the top right of each chunk
#This first chucnk is to load the neccesary libraries with install command if not yet installed
```{r}
#install.packages("ncdf4")
#install.packages("ggplot2")
#install.packages("dplyr")
#install.packages("osfr")
library(ncdf4)
library(ggplot2)
library(dplyr)
library(osfr)
```

# This next chunk is used to acces the temprature data file remotely from The Open Science Framework (OSF)
# You might need your own acces token, go to https://osf.io/settings/tokens/ to sign up and get one, then run:
# osf_auth("your_code")
#Next run this chunk to load tempraure data
```{r}
#This first line is to access the specific OSF project
osf_auth("FQ5MB6ryJmdnIPAbVl1WXC4G1gCYnmQ9QfsR4t5hVt80r7wsANTYoVWMThjvdSscfRagbI")
project <- osf_retrieve_node("au49d")

#Selecting the folder that contains the file with the "path =" command
temperature_files <- osf_ls_files(project, path = "temperature")

#Selecting the specific file (row) for download
temperature_file <- temperature_files[1, ]

#creating a temporary directory to download the file into
temp_dir <- tempdir()

#downloading the file into the temporary directory
osf_download(temperature_file, path = temp_dir, conflicts = "overwrite")
downloaded_file <- file.path(temp_dir, temperature_file$name)

#Open and extracting the data from the .nc file (NETcdf, and closing it to save memory
nc_file <- nc_open(downloaded_file)
unlink(downloaded_file)
```

# This next chunk is now used to dowload all the data from the Berkeley file we will be using for analysis
```{r}
# Extract latitude, longitude, time, and temperature data
lon <- ncvar_get(nc_file, "longitude")
lat <- ncvar_get(nc_file, "latitude")
time <- ncvar_get(nc_file, "time")
temperature_data <- ncvar_get(nc_file, "temperature")
climatology_data <- ncvar_get(nc_file, "climatology")

# Replace NaN values with NA
temperature_data[is.nan(temperature_data)] <- NA
climatology_data[is.nan(climatology_data)] <- NA

```


# In this next chunk we will be extracting all the temperature data for the grid
# Variables that needed to be altred are lat_indices and lon_indices (this was to specify specific coordinates of the grid).
# Another Variable that needs to be chnage is "ID" in the current_id line
# One more variable that can be changed is every time year is used but we did not alter this between runs

# This chunk needs to be rerun 12 times, changing the latitude/longitude and ID everytime before the chunk is ran
```{r}
# Get indices for the desired latitude and longitude ranges
# changed these for each Grid block
lat_indices <- which(lat >= 35 & lat <= 40)
lon_indices <- which(lon >= -140 & lon <= -110)

# Subset the temperature and climatology data for these specific indices
subset_temp <- temperature_data[lon_indices, lat_indices, ]
subset_climatology <- climatology_data[lon_indices, lat_indices, ]

# Verify that time is in years
years <- time

# Extract years 
# this can be changed to any date from 1850 to present, 
# the second value needs to be one year above desired last year
years_indices <- which(years >= 2000 & years <= 2024)

# These next few lines are to get the temperature anomalies and add them to a data frame
# Subset temperature data for the specified years
subset_temp_years <- subset_temp[,,years_indices]

# Ensure correct number of years
num_years <- length(years_indices)

# Initialize an array to store monthly data
monthly_data <- array(NA, dim = c(length(lon_indices), length(lat_indices), num_years, 12))

# Fill the array with subset_temp data
for (year in 1:num_years) {
  for (month in 1:12) {
    month_index <- (year - 1) * 12 + month
    if (month_index <= dim(subset_temp_years)[3]) {
      monthly_data[,,year,month] <- subset_temp_years[,,month_index]
    }
  }
}

# Calculate the mean temperature for each month over all years
mean_monthly_temp <- apply(monthly_data, c(3, 4), mean, na.rm = TRUE)

# Flatten the array and create a data frame
# change the value for years to earliest desired year
mean_monthly_temp_flattened <- as.vector(t(mean_monthly_temp))
months <- rep(1:12, num_years)
years <- rep(2000:(2000 + num_years - 1), each = 12)

# Create a data frame with months and temperatures anomalies
monthly_temp_df <- data.frame(Year = years, Month = months, Temperature = mean_monthly_temp_flattened)


# This next set of code is extract the baseline temperatures and make it possible to calculate absolute temperature      
# Calculate mean climatology temperature for each month
climatology_monthly_data <- array(NA, dim = c(length(lon_indices), length(lat_indices), 12))
for (month in 1:12) {
  climatology_monthly_data[,,month] <- apply(subset_climatology[,,month], c(1, 2), mean, na.rm = TRUE)
}

# Flatten the array and create a data frame for each month
mean_climatology_temp_flattened <- as.vector(climatology_monthly_data)
months <- rep(1:12, each = length(lon_indices) * length(lat_indices))

# Create a data frame with months and mean climatology temperatures
climatology_temp_df <- data.frame(
  Month = months,
  ClimatologyTemperature = mean_climatology_temp_flattened
)

# Remove NaN values if necessary
climatology_temp_df <- climatology_temp_df[!is.nan(climatology_temp_df$ClimatologyTemperature), ]

# Calculate mean climatology temperature for each month
climatology_monthly_means <- apply(subset_climatology, 3, mean, na.rm = TRUE)

# Create a data frame for the monthly averages
average_climatology_temp_df <- data.frame(
  Month = 1:12,
  AverageTemperature = climatology_monthly_means
)

# These next few lines merge the two data frames by month
# Merge the monthly temperature data frame with the average climatology temperature data frame
merged_df <- monthly_temp_df %>%
  left_join(average_climatology_temp_df, by = "Month")

# Sum the Temperature and AverageTemperature columns to get AbsoluteTemperature
# Change 'ID' to the ID of the grid section that is being done
current_id <- "ID"
current_df <- merged_df %>%
  mutate(!!current_id := Temperature + AverageTemperature)%>% # Calculating absolute temperature 
  select(-Temperature, -AverageTemperature)%>% # Cleaning up
  filter(Year <= 2023) # Filtering for last year

# Aligning by the Year, Month columns and adding the new ID to the dataframe each run
if (!exists("export_df")) {
  export_df <- current_df
} else {
  export_df <- full_join(export_df, current_df, by = c("Year", "Month"))
}
```


# Saving the final dataframe as a csv for export
write.csv(export_df, file = "Temperature.csv", row.names = FALSE)


