#Link for data
#https://berkeley-earth-temperature.s3.us-west-1.amazonaws.com/Global/Gridded/Complete_TAVG_LatLong1.nc


#load libraries
```{r}
library(ncdf4)
library(ggplot2)
library(dplyr)
```


#This chunk is used to get the first set of anomaly data
#Adjust Lat_indices and Lon_indices to set grid 
#Adjust years_indices to adjust for years, second value should be final year +1
#Change path to where file is on local computer(to big for git)
```{r}
# Open the NetCDF file
nc_file <- nc_open("C:\\Users\\wille\\Documents\\Pretorius_Willem\\Group-6-Owl-Migration\\Complete_TAVG_LatLong1.nc")

# Extract latitude, longitude, time, and temperature data
lon <- ncvar_get(nc_file, "longitude")
lat <- ncvar_get(nc_file, "latitude")
time <- ncvar_get(nc_file, "time")
temperature_data <- ncvar_get(nc_file, "temperature")

# Close the NetCDF file
nc_close(nc_file)

# Replace NaN values with NA
temperature_data[is.nan(temperature_data)] <- NA

# Get indices for the desired latitude and longitude ranges
lat_indices <- which(lat >= 30 & lat <= 50)
lon_indices <- which(lon >= -130 & lon <= -60)

# Subset the temperature data for these indices
subset_temp <- temperature_data[lon_indices, lat_indices, ]

# Verify that time is in years and use as is
years <- time

# Extract years 
years_indices <- which(years >= 2000 & years <= 2021)
print(length(years_indices))  # Check the length of years_indices

# Subset temperature data for these years
subset_temp_years <- subset_temp[,,years_indices]
print(dim(subset_temp_years))  # Verify the dimensions of the subset data

# Ensure correct number of years
num_years <- length(years_indices)

# Initialize the array to store monthly data
monthly_data <- array(NA, dim = c(length(lon_indices), length(lat_indices), num_years, 12))

# Fill the array with subset_temp data
for (year in 1:num_years) {
  for (month in 1:12) {
    month_index <- (year - 1) * 12 + month
    if (month_index <= dim(subset_temp_years)[3]) {
      monthly_data[,,year,month] <- subset_temp_years[,,month_index]
    }
  }
}

# Calculate the mean temperature for each month over all years
mean_monthly_temp <- apply(monthly_data, c(3, 4), mean, na.rm = TRUE)

# Flatten the array and create a data frame
mean_monthly_temp_flattened <- as.vector(t(mean_monthly_temp))
months <- rep(1:12, num_years)
years <- rep(2000:(2000 + num_years - 1), each = 12)

# Check lengths
print(length(mean_monthly_temp_flattened))
print(length(months))
print(length(years))

# Create a data frame with months and mean temperatures
monthly_temp_df <- data.frame(Year = years, Month = months, Temperature = mean_monthly_temp_flattened)

```
#This chunk is used to get the baseline data (1951 to 1980) from this data set
#Adjust la_indices and lon_indices to desired gridsize
```{r}
# Open the NetCDF file
nc_file <- nc_open("C:\\Users\\wille\\Documents\\Pretorius_Willem\\Group-6-Owl-Migration\\Complete_TAVG_LatLong1.nc")

# Extract latitude, longitude, and climatology data
lon <- ncvar_get(nc_file, "longitude")
lat <- ncvar_get(nc_file, "latitude")
climatology_data <- ncvar_get(nc_file, "climatology")

# Close the NetCDF file
nc_close(nc_file)

# Replace NaN values with NA
climatology_data[is.nan(climatology_data)] <- NA

# Get indices for the desired latitude and longitude ranges
lat_indices <- which(lat >= 30 & lat <= 50)
lon_indices <- which(lon >= -130 & lon <= -60)

# Subset the climatology data for these indices
subset_climatology <- climatology_data[lon_indices, lat_indices, ]

# Verify the dimensions and check for NaNs
print(dim(subset_climatology))
print(summary(as.vector(subset_climatology)))
      

# Calculate mean climatology temperature for each month
climatology_monthly_data <- array(NA, dim = c(length(lon_indices), length(lat_indices), 12))
for (month in 1:12) {
  climatology_monthly_data[,,month] <- apply(subset_climatology[,,month], c(1, 2), mean, na.rm = TRUE)
}

# Flatten the array and create a data frame for each month
mean_climatology_temp_flattened <- as.vector(climatology_monthly_data)
months <- rep(1:12, each = length(lon_indices) * length(lat_indices))

# Create a data frame with months and mean climatology temperatures
climatology_temp_df <- data.frame(
  Month = months,
  ClimatologyTemperature = mean_climatology_temp_flattened
)

# Remove NaN values if necessary
climatology_temp_df <- climatology_temp_df[!is.nan(climatology_temp_df$ClimatologyTemperature), ]

# Verify data
summary(climatology_temp_df)
```

#This is just to visualize the basline tempratures 
```{r}
# Use ggplot2 to create a plot
ggplot(climatology_temp_df, aes(x = factor(Month), y = ClimatologyTemperature, fill = ClimatologyTemperature)) +
  geom_boxplot() +
  labs(title = "Monthly Climatology Temperature (1951-1980)",
       x = "Month",
       y = "Temperature (°C)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

#Averging the baseline data into a dataframe
```{r}
# Calculate mean climatology temperature for each month
climatology_monthly_means <- apply(subset_climatology, 3, mean, na.rm = TRUE)

# Create a data frame for the monthly averages
average_climatology_temp_df <- data.frame(
  Month = 1:12,
  AverageTemperature = climatology_monthly_means
)

# Verify values
print(average_climatology_temp_df)
```
#Merged the Anomoly and baseline data
#Also cleasn the data
#In cleaned_df adjust year to last year in range
```{r}
# Merge the monthly temperature data frame with the average climatology temperature data frame
merged_df <- monthly_temp_df %>%
  left_join(average_climatology_temp_df, by = "Month")

# Verify the merged data frame
print(head(merged_df))

# Sum the Temperature and AverageTemperature columns to get AbsoluteTemperature
merged_df <- merged_df %>%
  mutate(AbsoluteTemperature = Temperature + AverageTemperature)

# Filter out rows where Year is greater than 2020
cleaned_df <- merged_df %>%
  filter(Year <= 2020)
```

#Create a plot from data to visualize trends
#Adjust plot_custom_months for specific months
```{r}
# Function to create plot for specified months
plot_custom_months <- function(months) {
  # Filter dataframe for specified months
  filtered_df <- cleaned_df %>% 
    filter(Month %in% months)
  
  # Create plot
  ggplot(filtered_df, aes(x = Year, y = AbsoluteTemperature, color = factor(Month))) +
    geom_line() +
    geom_smooth(method = "lm", se = FALSE, aes(group = Month, color = factor(Month)), linetype = "dashed") +
    labs(title = "Absolute Temperature for Selected Months",
         x = "Year",
         y = "Absolute Temperature (°C)",
         color = "Month") +
    theme_minimal()
}

# pick months
plot_custom_months(c(2, 3, 4))
```

#Slope and summary for speciefied months
#Adjust month ranges as desired
```{r}
# Function to calculate slopes for specified months
calculate_slopes <- function(months) {
  slopes <- cleaned_df %>%
    filter(Month %in% months) %>%
    group_by(Month) %>%
    summarize(slope = coef(lm(AbsoluteTemperature ~ Year))[2])
  
  return(slopes)
}

# Calculate slopes 
slopes <- calculate_slopes(c(1: 12))
print(slopes)

# Function to generate summary statistics for specified months
generate_monthly_summary <- function(months) {
  summary_df <- cleaned_df %>%
    filter(Month %in% months) %>%
    group_by(Month) %>%
    summarize(
      Min = min(AbsoluteTemperature, na.rm = TRUE),
      FirstQuartile = quantile(AbsoluteTemperature, 0.25, na.rm = TRUE),
      Median = median(AbsoluteTemperature, na.rm = TRUE),
      Mean = mean(AbsoluteTemperature, na.rm = TRUE),
      ThirdQuartile = quantile(AbsoluteTemperature, 0.75, na.rm = TRUE),
      Max = max(AbsoluteTemperature, na.rm = TRUE),
      StdDev = sd(AbsoluteTemperature, na.rm = TRUE)
    )
  
  return(summary_df)
}

# Generate summary
monthly_summary <- generate_monthly_summary(1 : 12)
print(monthly_summary)
```

#Save it as a csv
```{r}
# Select only the Year, Month, and AbsoluteTemperature columns
export_df <- cleaned_df %>%
  select(Year, Month, AbsoluteTemperature)
```
```{r}

# Save the selected columns to a CSV file
write.csv(export_df, "Absolute_temperature.csv", row.names = FALSE)
```

